An Effective Model of HL-60 Differentiation

### 2017-05-16 Paper_v2.tex
New Fig 9 edits

We hypothesized that Vav1 is a member of an ATRA-inducible complex which propels sustained MAPK activation, arrest and differentiation.
Initially, ATRA-induced Vav1 expression drives increased association between Vav1 and Raf.
This increased interaction facilitates phosphorylation and activation of Raf by pre-bound Akt and/or CK2 at S621 or perhaps S259.
Constitutively bound 14-3-3 may also stabilize the S621 phosphorylation, modulate the activity and/or up-regulate autophosphorylation.
Activated Raf can then drive ERK activation, which in turn closes the positive feedback loop by activating Raf transcription factors
e.g., Sp1 and/or STAT1 \cite{Kim2009,Milanini-Mongiat2002,Zhang2004,Li2006}.
We tested this working hypothesis using mathematical modeling.
The model recapitulated both ATRA time-course data as well as the GW5074 inhibitor effects.
This suggested the proposed Raf-Vav1 architecture was at least consistent with the experimental studies.
Further, analysis of the Raf-Vav1 model identified bistability in phosphorylated ERK levels.
Thus, two possible MAPK activation branches were possible for experimentally testable ATRA values.
The analysis also suggested the ATRA-induced Raf-Vav1 architecture could be locked into a
sustained signaling mode (high phosphorylated ERK) even in the absence of a ATRA signal.
This locked-in property could give rise to an ATRA-induction memory.
We validated the treatment memory property predicted by the Raf-Vav1 circuit experimentally using ATRA-washout experiments.
ERK phosphorylation levels remained high for more then 96 hr after ATRA was removed.
Previous studies demonstrated that HL-60 cells possessed an inheritable memory of ATRA stimulus \cite{YEN1984}.
Although the active state was self-sustaining, the inactive state demonstrated considerable robustness to perturbation.
For example, we found that 50x overexpression of Raf was required to reliably lock MAPK into the activated state, while small perturbations had almost no effect on phosphorylated ERK levels over the entire ensemble.
CD38 expression correlated with the phosphorylated ERK, suggesting its involvement in the signaling complex.
Our computational and experimental results showed that positive feedback,
through ERK-dependent Raf expression, could sustain MAPK signaling through many division cycles.
Such molecular mechanisms could underly aspects of cellular memory associated to consecutive ATRA treatments.

@@

\subsubsection*{Effective gene expression model equations.}
The ATRA differentiation model was encoded as a system of differential algebraic equations (DAEs) which described signaling and gene expression processes.
We modeled transcription and translation as Ordinary Differential Equations (ODEs), while signaling processes were assumed to quickly equilibrate and were
treated as a pseudo steady state system of algebraic equations.  
We decomposed the ATRA-induced differentiation program into three modules;
a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RAR/RXR (activated Trigger) signals;
a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals; and
a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors.
The output of the signal initiation module was the input to the gene expression model.
For each gene $j=1,2,\dots,\mathcal{G}$, we modeled both the mRNA ($m_{j}$), protein ($p_{j}$) and signaling species abundance:
\begin{eqnarray}
	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\theta_{m,j}\right)m_{j}+\lambda_{j}\\
	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\theta_{p,j}\right)p_{j}\\\label{eqn-signaling-balances}
	\mathbf{g}\left(p_{1},\hdots,p_{\mathcal{G}},\kappa\right) &=& \mathbf{0}
\end{eqnarray}
where signaling species abundance was governed by the non-linear algebraic equations $\mathbf{g}\left(p_{1},\hdots,p_{\mathcal{G}},\kappa\right) = \mathbf{0}$.
The model parameter vector is denoted by $\kappa$.
The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
the terms $\theta_{m,j}$ and $\theta_{p,j}$ denote first-order degradation constants for mRNA and protein, respectively.
The specific transcription rate $r_{T,j}$ was modeled as the product of a kinetic term $\bar{r}_{T,j}$ and a control term $u_{j}$ which described how the
abundance of transcription factors, or other regulators influenced the expression of gene $j$.
The kinetic transcription term $\bar{r}_{T,j}$ was modeled as:
\begin{equation}
	\bar{r}_{T,j} = V^{max}_{T}\left(\frac{L_{T,o}}{L_{T,j}}\right)\left(\frac{G_{j}}{K_{T}+G_{j}}\right)
\end{equation}where the maximum gene expression rate $V^{max}_{T}$ was defined as the product of a characteristic transcription rate constant ($k_{T}$)
and the abundance of RNA polymerase ($R_{1}$), $V^{max}_{T} = k_{T}\left(R_{1}\right)$.
The $\left(L_{T,o}/L_{T,j}\right)$ term denotes the ratio of transcription read lengths; $L_{T,o}$ represents a characteristic gene length, while $L_{T,j}$ denotes the length of gene $j$.
Thus, the ratio $\left(L_{T,o}/L_{T,j}\right)$ is a gene specific correction to the characteristic transcription rate $V^{max}_{T}$.
Lastly, the $\lambda_{j}$ term denotes the constitutive rate of expression of gene $j$.

The gene expression control term $0\leq u_{j}\leq 1$ depended upon the combination of factors which influenced the expression of gene $j$.
If the expression of gene $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$u_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on the expression of gene $j$,
and $\mathcal{I}_{j}\left(\cdot\right)$ denotes an integration rule which combines the individual regulatory inputs for gene $j$ into a single control term.
In this study, the integration rule governing gene expression was the weighted fraction of promoter configurations that resulted in gene expression \cite{Moon:2012ab}:
\begin{equation}
	{u}_{j} = \frac{W_{R_{1},j}+\displaystyle\sum_{n}W_{nj}f_{nj}}{1+W_{R_{1},j}+\displaystyle\sum_{d}W_{dj}f_{dj}}
\end{equation}
The numerator, the weighted sum (with weights $W_{nj}$) of promoter configurations leading to gene expression, was normalized by all possible promoter configurations.
The likelihood of each configuration was quantified by the transfer function $f_{nj}$ (which we modeled using Hill like functions), while
the lead term in the numerator $W_{R_{1},j}$ denotes the weight of constitutive expression for gene $j$.
Given this formulation, the rate of constitutive expression was then given by:
\begin{equation}
	\lambda_{j} = \bar{r}_{T,j}\left(\frac{W_{R_{1},j}}{1+W_{R_{1},j}}\right)
\end{equation}
If a gene expression process had no modifying factors, $u_{j}=1$.
Lastly, the specific translation rate was modeled as:
\begin{equation}
	r_{X,j} = V^{max}_{X}\left(\frac{L_{X,o}}{L_{X,j}}\right)\left(\frac{m_{j}}{K_{X}+m_{j}}\right)
\end{equation}
where $V^{max}_{X}$ denotes a characteristic maximum translation rate estimated from literature, and $K_{X}$ denotes a translation saturation constant.
The characteristic maximum translation rate was defined as the product of a characteristic translation rate constant ($k_{X}$)
and the Ribosome abundance ($R_{2}$), $V^{max}_{X} = k_{X}\left(R_{2}\right)$. As was the case for transcription,
we corrected the characteristic translation rate by the ratio of the length of a characteristic transcription normalized by the length of transcript $j$.

@@

\begin{figure}[!t]\centering
\includegraphics[width=0.85\textwidth]{./figs/Fig-4-GRN-Simulations.pdf}
\caption{Model simulation of the HL-60 gene expression program following exposure to 1$\mu$M ATRA at t = 0 hr.
A: CD38 and CD11b expression versus time following ATRA exposure at time t = 0 hr.
B: Gene expression at t = 48 hr following ATRA exposure.
Experimental data in panels A and B were reproduced from Jensen et al. \cite{Jensen:2015aa}.}\label{fig:model-grn-simulations}
\end{figure}

\begin{figure}[!t]\centering
\includegraphics[width=0.85\textwidth]{./figs/Fig-5-GRN-CellCycle-Predictions.pdf}
\caption{Model simulation of HL-60 cell-cycle arrest following exposure to 1$\mu$M ATRA at t = 0 hr.
A: Predicted p21 and E2F expression levels for the best parameter set following ATRA exposure at time t = 0 hr.
B: Estimated fraction of HL-60 cells in G0 arrest following ATRA exposure at time t = 0 hr.
The gray region denotes the 95\% confidence estimate of the polynomial model.
Experimental data in panel B was reproduced from Jensen et al. \cite{Jensen:2015aa}.}\label{fig:model-cell-cycle-simulations}
\end{figure}

@@

\begin{figure}[!t]
\includegraphics[width=0.9\textwidth]{./figs/Fig-9-ATRA-Schematic.pdf}
\caption{Schematic of the hypothetical principal pathways in the ATRA-induced
signaling that results in cell differentiation in the HL-60 myeloid leukemia model
\cite{wallace2016rrd,congleton2012src,congleton2014phosphorylation,Shen:2009aa,Bunaciu:2015aa,shen2008c}.
It is based on modules and feedback loops. There are three main arms (shown top to bottom):
1. Direct ATRA targeting of RAREs in genes such as CD38 or BLR1;
2. Formation of a signalsome that has a
regulatory module that includes Vav (a GEF), CBL and SLP-76 (adaptors), and Lyn (a SFK) that
regulates a Raf/Mek/Erk axis that incorporates Erk to Raf feedback, where the regulators are
modulated by AhR and CD38 receptors; and 3. Direct ATRA targeted up regulation of CDKI to
control RB hypophosphorylation. The Raf/Mek/Erk axis is embedded in the signalsome and
subject to modulation by the regulators. The output of the signalsome is discharge of the Raf
from the cytosol to the nucleus where it binds (hyper)phospho-RB and other targets, including
NFATc3, which enables activation of the RA bound RAR/RXR poised on the BLR1 promoter,
and also GSK3, phosphorylation of which relieves its inhibitory effect on RAR$\alpha$. CDKI directed
hypophosphorylation of RB releases Raf sequestered by RB to go to NFATc3, GSK3, and other
targets. A significant consequence of the nuclear RAF is ergo ultimately to hyperactivate
transcriptional activation by RAR$\alpha$ to drive differentiation.
It might be noted that this general proposed model provides a mechanistic rationalization for
why cell cycle arrest is historically oft times perceived as a feature predisposing phenotypic
maturation.}
\end{figure}

>>>>

We hypothesized that Vav1 is a member of an ATRA-inducible signalsome complex which propels sustained MAPK activation, arrest and differentiation (shown schematically in Fig. \ref{fig:signalsome-schematic}).
Initially, ATRA-induced Vav1 expression drives increased association between Vav1 and Raf.
This increased interaction facilitates phosphorylation and activation of Raf by pre-bound Akt and/or CK2 at S621 or perhaps S259.
Constitutively bound 14-3-3 may also stabilize the S621 phosphorylation, modulate the activity and/or up-regulate autophosphorylation.
Activated Raf can then drive ERK activation, which in turn closes the positive feedback loop by activating Raf transcription factors
e.g., Sp1 and/or STAT1 \cite{Kim2009,Milanini-Mongiat2002,Zhang2004,Li2006}.
We tested this working hypothesis using mathematical modeling.
The model recapitulated both ATRA time-course data as well as the GW5074 inhibitor effects.
This suggested the proposed Raf-Vav1 architecture was at least consistent with the experimental studies.
Further, analysis of the Raf-Vav1 model identified bistability in phosphorylated ERK levels.
Thus, two possible MAPK activation branches were possible for experimentally testable ATRA values.
The analysis also suggested the ATRA-induced Raf-Vav1 architecture could be locked into a
sustained signaling mode (high phosphorylated ERK) even in the absence of a ATRA signal.
This locked-in property could give rise to an ATRA-induction memory.
We validated the treatment memory property predicted by the Raf-Vav1 circuit experimentally using ATRA-washout experiments.
ERK phosphorylation levels remained high for more then 96 hr after ATRA was removed.
Previous studies demonstrated that HL-60 cells possessed an inheritable memory of ATRA stimulus \cite{YEN1984}.
Although the active state was self-sustaining, the inactive state demonstrated considerable robustness to perturbation.
For example, we found that 50x overexpression of Raf was required to reliably lock MAPK into the activated state, while small perturbations had almost no effect on phosphorylated ERK levels over the entire ensemble.
CD38 expression correlated with the phosphorylated ERK, suggesting its involvement in the signaling complex.
Our computational and experimental results showed that positive feedback,
through ERK-dependent Raf expression, could sustain MAPK signaling through many division cycles.
Such molecular mechanisms could underly aspects of cellular memory associated to consecutive ATRA treatments.

@@

\subsubsection*{Effective gene expression model equations.}
The ATRA differentiation model was encoded as a system of differential algebraic equations (DAEs) which described both signaling and gene expression processes.
We modeled transcription and translation as Ordinary Differential Equations (ODEs), while signaling processes were assumed to quickly equilibrate and were
treated as a pseudo steady state system of algebraic equations.
We decomposed the ATRA-induced differentiation program into three modules;
a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RAR/RXR (activated Trigger) signals;
a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals; and
a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors.
The output of the signal initiation module was the input to the gene expression model.
For each gene $j=1,2,\dots,\mathcal{G}$, we modeled both the mRNA ($m_{j}$), protein ($p_{j}$) and signaling species abundance:
\begin{eqnarray}
	\frac{dm_{j}}{dt} &=& r_{T,j} - \left(\mu+\theta_{m,j}\right)m_{j}+\lambda_{j}\\
	\frac{dp_{j}}{dt} &=& r_{X,j} - \left(\mu+\theta_{p,j}\right)p_{j}\\\label{eqn-signaling-balances}
	\mathbf{g}\left(p_{1},\hdots,p_{\mathcal{G}},\kappa\right) &=& \mathbf{0}
\end{eqnarray}
where signaling species abundance was governed by the non-linear algebraic equations $\mathbf{g}\left(p_{1},\hdots,p_{\mathcal{G}},\kappa\right) = \mathbf{0}$.
The model parameter vector is denoted by $\kappa$.
The terms $r_{T,j}$ and $r_{X,j}$ denote the specific rates of transcription, and translation while
the terms $\theta_{m,j}$ and $\theta_{p,j}$ denote first-order degradation constants for mRNA and protein, respectively.
The specific transcription rate $r_{T,j}$ was modeled as the product of a kinetic term $\bar{r}_{T,j}$ and a control term $u_{j}$ which described how the
abundance of transcription factors, or other regulators influenced the expression of gene $j$.

The gene expression control term $0\leq u_{j}\leq 1$ depended upon the combination of factors which influenced the expression of gene $j$.
If the expression of gene $j$ was influenced by $1,\dots,m$ factors, we modeled this relationship as
$u_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on the expression of gene $j$,
and $\mathcal{I}_{j}\left(\cdot\right)$ denotes an integration rule which combines the individual regulatory inputs for gene $j$ into a single control term.
In this study, the integration rule governing gene expression was the weighted fraction of promoter configurations that resulted in gene expression \cite{Moon:2012ab}:
\begin{equation}
	{u}_{j} = \frac{W_{R_{1},j}+\displaystyle\sum_{n}W_{nj}f_{nj}}{1+W_{R_{1},j}+\displaystyle\sum_{d}W_{dj}f_{dj}}
\end{equation}
The numerator, the weighted sum (with weights $W_{nj}$) of promoter configurations leading to gene expression,
was normalized by all possible promoter configurations (denominator).
The likelihood of each configuration was quantified by the transfer function $f_{nj}$ (which we modeled using Hill functions), while
the lead term in the numerator $W_{R_{1},j}$ denotes the weight of constitutive expression for gene $j$.
Given the formulation of the control law, the $\lambda_{j}$ term (which denotes the constitutive rate of expression of gene $j$) was given by:
\begin{equation}
	\lambda_{j} = \bar{r}_{T,j}\left(\frac{W_{R_{1},j}}{1+W_{R_{1},j}}\right)
\end{equation}

The kinetic transcription term $\bar{r}_{T,j}$ was modeled as:
\begin{equation}
	\bar{r}_{T,j} = V^{max}_{T}\left(\frac{L_{T,o}}{L_{T,j}}\right)\left(\frac{G_{j}}{K_{T}+G_{j}}\right)
\end{equation}where the maximum gene expression rate $V^{max}_{T}$ was defined as the product of a characteristic transcription rate constant ($k_{T}$)
and the abundance of RNA polymerase ($R_{1}$), $V^{max}_{T} = k_{T}\left(R_{1}\right)$.
The $\left(L_{T,o}/L_{T,j}\right)$ term denotes the ratio of transcription read lengths; $L_{T,o}$ represents a characteristic gene length, while $L_{T,j}$ denotes the length of gene $j$.
Thus, the ratio $\left(L_{T,o}/L_{T,j}\right)$ is a gene specific correction to the characteristic transcription rate $V^{max}_{T}$.
If a gene expression process had no modifying factors, $u_{j}=1$.
Lastly, the specific translation rate was modeled as:
\begin{equation}
	r_{X,j} = V^{max}_{X}\left(\frac{L_{X,o}}{L_{X,j}}\right)\left(\frac{m_{j}}{K_{X}+m_{j}}\right)
\end{equation}
where $V^{max}_{X}$ denotes a characteristic maximum translation rate estimated from literature, and $K_{X}$ denotes a translation saturation constant.
The characteristic maximum translation rate was defined as the product of a characteristic translation rate constant ($k_{X}$)
and the Ribosome abundance ($R_{2}$), $V^{max}_{X} = k_{X}\left(R_{2}\right)$. As was the case for transcription,
we corrected the characteristic translation rate by the ratio of the length of a characteristic transcription normalized by the length of transcript $j$.
The sequence lengths used in this study are given in Table \ref{tbl:TF-sequences}.

@@

\begin{figure}[!t]\centering
\includegraphics[width=0.70\textwidth]{./figs/Fig-4-GRN-Simulations.pdf}
\caption{Model simulation of the HL-60 gene expression program following exposure to 1$\mu$M ATRA at t = 0 hr.
A: Scaled CD38 and CD11b expression versus time following ATRA exposure at time t = 0 hr.
B: Scaled Gene expression at t = 48 hr following ATRA exposure.
Gene expression was normalized to expression in the absence of ATRA.
Experimental data in panels A and B were reproduced from Jensen et al. \cite{Jensen:2015aa}.
Model simulations were conducted using the ten best parameter sets collected during model identification.
Solid lines (or bars) denote the mean model performance, while the shaded region (or error bars) denote the 95\% confidence
interval calculated over the parameter ensemble.}\label{fig:model-grn-simulations}
\end{figure}

\begin{figure}[!t]\centering
\includegraphics[width=0.85\textwidth]{./figs/Fig-5-GRN-CellCycle-Predictions.pdf}
\caption{Model simulation of HL-60 cell-cycle arrest following exposure to 1$\mu$M ATRA at t = 0 hr.
A: Predicted p21 and E2F expression levels for the best parameter set following ATRA exposure at time t = 0 hr.
B: Estimated fraction of HL-60 cells in G0 arrest following ATRA exposure at time t = 0 hr.
Solid lines (or bars) denote the mean model performance, while the shaded region (or error bars) denotes the 95\% confidence estimate of the polynomial model.
Experimental data in panel B was reproduced from Jensen et al. \cite{Jensen:2015aa}.}\label{fig:model-cell-cycle-simulations}
\end{figure}

@@

\begin{figure}[!t]
\includegraphics[width=0.9\textwidth]{./figs/Fig-9-ATRA-Schematic.pdf}
\caption{Schematic of the hypothetical principal pathways implicated in ATRA-induced differentiation of HL-60 myeloid leukemia cells
\cite{wallace2016rrd,congleton2012src,congleton2014phosphorylation,Shen:2009aa,Bunaciu:2015aa,shen2008c}.
There are three main arms (shown top to bottom):
(a) Direct ATRA targeting of RAREs in genes such as CD38 or BLR1;
(b) Formation of a signalsome that has a regulatory module that includes Vav (a guanine nucleotide exchange factor),
CBL and SLP-76 (adaptors), and Lyn (a Src family kinase) that
regulates a Raf/Mek/Erk axis that incorporates Erk to Raf feedback, where the regulators are modulated by AhR and CD38 receptors; and
(c) Direct ATRA upregulation of CDKIs to control RB hypophosphorylation.
The Raf/Mek/Erk axis is embedded in the signalsome and
subject to modulation by the regulators. The signalsome discharges Raf
from the cytosol to the nucleus where it binds (hyper)phospho-RB and other targets, including
NFATc3, which enables activation of the ATRA bound RAR/RXR poised on the BLR1 promoter,
and GSK3, phosphorylation of which relieves the inhibitory effect on RAR$\alpha$.
CDKI directed hypophosphorylation of RB releases Raf sequestered by RB to go to NFATc3, GSK3, and other targets.
A significant consequence of the nuclear Raf is ergo ultimately to hyperactivate
transcriptional activation by RAR$\alpha$ to drive differentiation.
It might be noted that this general proposed model provides a mechanistic rationalization for
why cell cycle arrest is historically oft times perceived as a feature predisposing phenotypic
maturation.}\label{fig:signalsome-schematic}
\end{figure}
